Simplex: An individual-based platform for ecological modeling

### 2015-11-03 paper/PrePrint.Rmd

Here, I present a detailed description for an ecological IBM platform I refer to as "Simplex". The Simplex platform simulates a broad range of ecological conditions, records information from individuals to the ecosystem level, and provides computing code for analyzing output. Simplex incorporates fluid and non-fluid dynamics with stochastic and selection-based changes within and among populations. Simplex also integrates nutrient-limited growth and physiology. I constructed Simplex with the aim of simulating ecologically complex scenarios, from which the results could be used to test (*i*) predictions of ecological theory and (*ii*) the general influence of constraints and processes on populations, communities, and ecosystems. I provide an explanation of how Simplex works, the data it quantifies and tracks, the theories and principles that Simplex integrates, and the analyses that can be conducted using the open source code.

@@

**Model description following ODD Protocol**  
The ODD protocol is standard for describing individual-based models (Grimm et al. 2006). ODD stands for Overview, Design concepts, and Details. Here, I provide an abbreviated explanation of the Simplex ODD protocol. The full ODD protocol can be found on the public GitHub repository at https://github.com/LennonLab/simplex. Likewise, to save space, I refer the reader to the README.md file on the public Simplex GitHub.com repository (https://github.com/LennonLab/simplex/blob/master/README.md) for descriptions of source code files and directories.

**Purpose**  
The purpose of Simplex is to explicitly simulate life history and the evolution of traits among individual organisms, the assembly of ecological communities, and the structure and function of ecosystems in a spatially explicit environment under stochastic conditions. The proximate goal of Simplex is to generate high degrees of variation in the assembly and structure of populations and communities by assembling many different models from random combinations of state-variables and processes. The ultimate goal of Simplex is to provide a simulation-based platform for examining conditions under which processes and constraints have a robust influence on eco-evolutionary dynamics, patterns of biodiversity, and ecosystem processes.

@@

*Individual organisms*--Individuals are distinguished by collections of elements within lists. Individuals undergo changes when randomly sampled from lists. Each specific position in the list corresponds to the same individual. For example, in simulating growth, a Simplex model chooses an individual from the list of cell quotas (the probability of reproducing is determined by endogenous resources). The first position in this list as well in all other lists of individual attributes corresponds to the same individual. Lists of attributes for individual organisms include the time each individual spends in the system, species ID, individual ID, 2D spatial location, endogenous levels of resources (resource specific cell quotas), individual-level metabolic maintenance cost, individual-level maximum dispersal rate, parental lineage, individual resource use efficiency for each resource.

*Species*--Each species is characterized by the individuals that share a common set of traits, such as maximum growth rate, metabolic maintenance cost. Species information is stored in Python dictionaries. In this way, if Simplex requires the species ID of an individual it will access the species ID list where each element corresponds to a specific individual. But, if Simplex requires the maximum growth rate for an individual, then it finds the species ID and then uses that to accesses the Python dictionary for maximum specific growth rates of species. The species-level information that are stored include metabolic maintenance cost, maximum dispersal rate, maximum theoretical growth rate, and the resource use efficiency for each resource.

*Resource particles*--Individual resource particles are distinguished by collections of elements within lists. Each specific position in the list corresponds to the same resource particle. The types of information stored about each resource particle include time in the system, particle ID, 2D spatial location, the type of resource particle, the size of the particle.

*Inert tracers particles*--These are objects that move/flow into and through the environment. They only interact with physical barriers. The use of tracers particles allows for attributes of a system that flows or physically turns over to be quantified. Like other individual-level objects in Simplex models, tracers are distinguished by collections of elements within lists. Each specific position in the list corresponds to the same resource particle. Information stored about each resource particle includes time spent in the system, 2D spatial location and particle ID.

@@

**System level state variables**  
Each run of Simplex begins with random choices for the values of width (5 to 100), height (5 to 100), basal flow rate (0.0 to 1.0), the number, size, and location of physical barriers, the number and direction of environmental gradients, rate of stochastic disturbance, rate of fluctuation in basal flow rate, and degrees of fluctuation and syncronized flow of individual particles, ranging from completely in-sync or completely out of sync.

**Spatial and temporal scales**  
\setlength\parindent{36pt}
The two general aspects of scale are grain (i.e., resolution) and extent (e.g. total area). The environment of Simplex models is two dimensional and can vary along each axis from 5 to 100 discrete units. This makes for a potential total extent of 25 to 10,000 discrete patches, each with a grain of 1 square unit. All particles move in decimal units, meaning that individual particles can  practically infinite locations within patches and likewise. Extent of time in Simplex models refers to residence time, i.e., the average amount of time that individual particles spend in the system. Residence time for individual particles (organisms, resources, tracers) can vary across five orders of magnitude.

Grain is the smallest unit over which change can happen. For example, the original ODD documentation states “One time step represents one year and simulations were run for 100 years. One grid cell represents 1 ha and the model landscape comprised 1,000 x 1,000 ha; i.e., 10,000 square kilometers”. In contrast, the smallest grain achievable by Simplex is determined by slowest rate at which individuals can undergo BIDE (birth, immigration, death, emigration) processes. For example, under high residence times, an individual can move across 0.00001% of the x or y axis in one time step. Under low residence times, an individual can move across 10% or greater of the x or y axis in one time step.

@@

*Core simulation process*. Simplex models begin simulation immediately after assembly from random combinations of state-variables and processes. Instead of operating by definitive time steps (i.e. days, generations), Simplex models advance turnover of the environmental matrix according to the initial rate of flow. If the initial rate of flow is 1.0, then the environmental matrix and inert particles would flow 1.0 units of distance. After each iteration of flow, each individual is given the chance to consume, grow, reproduce, starve, die, and to disperse towards resources and environmental optima.

*Simulated life history*.\setlength\parindent{36pt} Simplex models simulate growth, reproduction, and death via weighted random sampling. This simulate the partly probabilistic and partly deterministic nature of environmental filtering and individual-level interactions.

Inflow/Entrance. Resources and individuals enter from any point in the environment. Species identities of inflowing propagules are chosen at random from a log-series distribution, which often approximates the distribution of abundance among species in ecological communities (see Hubbell 2001). Along with a species ID and species-specific maximum rates of resource uptake, active dispersal, resource efficiencies, cell maintenance, and environmental optima, each propagule was given a unique ID and a multi-resource cell quota that represent the state of internal resources. The average of these cell quotas determine the probability of reproduction.

Dispersal. Individuals are allowed to actively move along environmental gradients (sometimes against the direction of environmenta flow) and towards their optima. A better match to one's environmental optima increases the chance of reproduction and the individual's ability to perform (consume, grow).

Consumption & growth. Sampled individuals consume resources according to their specific maximum rates of uptake and grow according to specific resource efficiencies and maintenance costs. Uptake increases the individual cell quotas and decreases ambient resources. Individual cell quotas are then decreased according to a specific physiological maintenance cost.

Reproduction. Reproduction in Simplex is currently limited to clonal reproduction with the possibility of mutation. Individuals reproduce with a probability determined by the combination of mean cell quota and the proportional match to the environmental optima. The cell quota of each resource is evenly divided between the individual and its daughter. The daughter is given a unique individual ID and the species ID of its mother, unless in the case of speciation, but is allowed small mutations in individual-level state variables.

@@

**Design concepts**  
*Ecological complexity*. Simplex assembles models from random combinations of constraints (state-variables) and processes to generate output data that allow the user to test the general influence of a particular state-variable, process, or combination thereof using univariate and multivariate tests.

*Nutrient limited growth*. All models assembled by Simplex employ the universal concept that individual growth and activity is fueled and limited by resources.

*Resource diversity & heterogeneity*. Simplex allows the user to explore the influence of the number and abundances of different resources on ecological diversity and ecosystem processes.

@@

These data are stored in file as R-formatted data.frames. These files can be directly imported into an R or Python environment.

@@

Droop (1968, 1983) gave a relationship between specific growth rate ($\mu$) and cell quota ($Q$): $$\mu = \mu'_{m} (1 - k_{q}/Q)$$
where $k_{q}$ is the minimum cell quota needed for life, also referred to as the subsistence quota. $\mu'$ is the

*Maintenance cost of Pirt*. Pirt (1965) states "The variation, with growth rate, of the yield of organism from the substrate used as energy source is attributed to consumption of energy at a constant rate for cell maintenance." He derived a relationship between the growth yield (biomass), the growth rate, and metabolic maintenance. Simplex models use Pirt's concept of a constant maintenance requirement. Simplex also draws from Pirt's simple relation for substrate use: $$use(total) = use(maintenance) + use(growth)$$

@@

The Simplex platform assembles individual-based models from random combinations of state-variables and processes, to explore a degrees of variation in ecological conditions and complexity. In this way, Simplex allows ecologists to examine whether the influence of particular variables should be robust to a wide range of ecological and environmental conditions. In addition to its unique modeling aspects, Simplex includes analysis code that shows users how to import and analyze simulated datasets, including variance partitioning and fitting of species abundance models.

Future developments of Simplex will have trophic dynamics including predator-prey, symbionts and mutualists, and producers and consumers. Individuals will also be attributed with additional behaviors (e.g., search patterns), sensing (e.g., of resources), and learning (e.g., predator avoidance). Additionally, individuals will also be attributed with genomes. Initially, these genomes will be represented by sets of sequences of A's, C's, T's, and G's representing genes that can be mutated as they're passed on through clonal reproduction, or mutated and recombined in the case of sexual haploid or diploid reproduction. A precursor to this has already been developed in Locey (2013), the code for which is available here: https://github.com/klocey/SYMENT.

By taking advantage of modern computing and version control and social coding tools (e.g., Python, R/RStudio, Knitr, Git, GitHub) Simplex is well-positioned to continue development as both a research and education/instructional tool for ecology. Future developments and released versions will include additional ecological dynamcis such as predator-prey, mutualism, and parasitism. Improvements to Simplex will also include an explicit stoichiometry, where individual resource particles have ratios of Carbon, Nitrogen, and Phosphorus, as well as degrees of biocomplexity. For example, given the chemical structure of phosphate and the size of a resource particle, the particle can be assigned *x* units of phosphorus as well as the biocomplexity value of phosphate (estimated as a form of Shannon's entropy).

Likewise, improvements and future versions of Simplex will provide increasing numbers of files for statistical analysis of Simplex output. These files will, in a sense, be able to be used as a complete analysis of some particular aspect of Simplex output (e.g., demographic, resource-related, spatial, multivariate, ecological model testing). These and all Simplex files are freely available and open source, allowing users to modify Simplex source code to suit their particular needs.

@@

1. Black AJ, McKane AJ (2012) Stochastic formulation of ecological models and their applications. Trends in Ecology and Evolution. 27:337-345.

2. DeAngelis DL, Gross LJ (1992) Individual-based Models and Approaches in Ecology: Populations, Communities, and Ecosystems. Chapman & Hall. New York.

3. Droop MR (1983) Twenty-five years of algal growth kinetics, a personal view. Botanica Marina. 26: 99-112.

4. Gotelli NJ, Entsminger GL (2001) Swap and fill algorithms in null model analysis: rethinking the knight’s tour. Oecologia. 129:281–291

5. Grimm V, Berger U, Bastiansen F, Eliassen S, Ginot V, Giske J, Goss-Custard J, Grand T, Heinz SK, Huse G, Huth A, Jepsen JU, Jørgensen C, Mooij WM, Mu¨ller B, Pe’er G, Pioub C, Railsback SF, Robbins AM, Robbins MM, Rossmanith E, Ru¨ger N, Strand E, Souissim S, Stillmane RA, Vabøg R, Visser U, DeAngelis DL (2006) A standard protocol for describing individual-based and agent-based models. Ecological Modeling. 198: 115-126.

6. Grimm V, Railsback SF (2005) Individual-based Modeling and Ecology. Princeton University Press, Princeton.

7. Hartl DL, Clark AG (1997) Principle of Population Genetics. Sinauer Associates, Inc. Publishers, Sunderland, MA.

8. Hubbell SP (2001) The unified neutral theory of biodiversity and biogeography. Princeton University Press, Princeton, NJ.

9. Leadbeater BSC (2006) The 'Droop equation' - Michael Droop and the legacy of the 'Cell-quota model' of phytoplankton growth. Protist. 157: 345-358.

10. Locey K (2013) Synthesized ecological molecular neutral theory. figshare. http://dx.doi.org/10.6084/m9.figshare.640105.

11. Locey KJ, Lennon JT. (2015) Scaling laws predict global microbial diversity. PeerJ PrePrints 3:e1808 https://dx.doi.org/10.7287/peerj.preprints.1451v1

12. Matthews RB, Gilbert NG, Roach A, Polhill JG, Gotts NM (2007) Agent-based land-use models: a review of applications. Landscape Ecology. 22: 1447-1459.

13. Pirt SJ (1965) The Maintenance Energy of Bacteria in Growing Cultures. Proceedings of the Royal Society of London. Series B, Biological Sciences, 163:224-231.

14. Rosindell J, Harmon LJ, Etienne RS (2015) Unifying ecology and macroevolution with individual-based theory. Ecology Letters. 18: 472-482.

15. Succi S (2001) The Lattice Boltzmann Equation: For Fluid Dynamics and Beyond. Oxford University Press. New York.

>>>>

Here, I present a description of an ecological IBM platform I refer to as "Simplex". This platform simulates a broad range of ecological conditions, records information from individuals to the ecosystem level, and provides R computing code for analyzing output. Simplex incorporates fluid and non-fluid dynamics with stochastic and selection-based changes within and among populations. Simplex also integrates nutrient-limited growth and physiology. I constructed Simplex with the aim of simulating ecologically complex scenarios, from which the results could be used to test (*i*) predictions of ecological theory and (*ii*) the general influence of constraints and processes on populations, communities, and ecosystems.

@@

**Model description following ODD Protocol**  
The overview, design concepts, and details protocol (ODD) is a standard for describing individual-based models (Grimm et al. 2006). Here, I provide an abbreviated version of the Simplex ODD protocol. The full Simplex ODD protocol can be found on the public GitHub repository at https://github.com/LennonLab/simplex. Likewise, to save space, I refer the reader to the README.md file on the public Simplex GitHub.com repository for descriptions of source code, files, and directories (https://github.com/LennonLab/simplex/blob/master/README.md).

**Purpose**  
The purpose of Simplex is to explicitly simulate life history, the evolution of traits, the assembly of ecological communities, and the structure and function of ecosystems in a spatially explicit environment under stochastic conditions, environmental filtering, and species sorting. The proximate goal of Simplex is to generate high degrees of variation in the assembly and structure of populations and communities by assembling many different models from random combinations of state-variables and processes. The ultimate goal of Simplex is to provide a simulation-based platform for examining conditions under which processes and constraints have a robust influence on eco-evolutionary dynamics, patterns of biodiversity, and ecosystem processes.

@@

*Individual organisms*--These are distinguished by the corresponding elements of several lists. For example, in simulating reproduction, a Simplex model chooses an individual at random from the list of cell quotas, where the probability of reproducing is determined by the levels of endogenous resources, i.e., cell quota. The first position in the list of cell quotas, and in all other lists of individual attributes, corresponds to the same individual, i.e.,:

```{r, eval=FALSE}
IndIDs = [1,  2, 33, 14]  
Quotas =  [0.1, 0.99, 0.14, 0.05]  
Xpos =   [45, 23, 456, 1]  
Ypos =   [765, 87, 21, 34]  
```

Above, the individual with the ID "1" has a cell quota of 0.1  and is located at position x=45, y=765. Note that in simplex, there are at least three resource quotas. In future versions, the quotas will more explicitly represent nitrogen, carbon, and phosphorus. Of course, each individual could also have been distinguished as its own list of attributes, e.g., Ind1 = [1, 0.1, 45, 765]. However, this would require a large lists of lists and could make quantifying some metrics (e.g., generating a frequency distribution of cell quotas, ages, or environmental optima) more computationally intensive than necessary.

Lists of attributes for individual organisms include the time each individual spends in the system, species ID, individual ID, 2D spatial location, endogenous levels of resources (resource specific cell quotas), individual-level metabolic maintenance cost, individual-level maximum dispersal rate, parental lineage, and individual resource use efficiency for each resource.

*Species*--Each species is characterized by the individuals that share a common set of traits, such as maximum growth rate, maximum dispersal rate, and metabolic maintenance cost. Species information is stored in dictionaries. In this way, if Simplex requires the species ID of an individual it will access the species ID list. But, if Simplex requires the maximum growth rate for an individual, it will first find the species ID and then use that ID to accesses the dictionary for maximum specific growth rates of species. Species-level data include metabolic maintenance cost, maximum dispersal rate, maximum theoretical growth rate, environmental optima, and the resource use efficiency for each resource.

```{r, eval=FALSE}
IndIDs = [1,  2, 33, 14]
spIDs = [12, 32, 11, 6]
MaxGrowthDict = {12: 0.9, 32: 0.5, 11: 0.8, 6: 0.4}
```

The individual with ID of 1 belongs to species 12, which has a theoretical maximum growth rate of 0.9. By storing species-level information in dictionaries, Simplex does not have to store redundant information in lists.

*Resource particles*--Like individual organisms, individual resource particles distinguished by the corresponding elements of several lists. Each list holds a specific type of information, including time spent in the system, particle ID, 2D spatial location, the type of resource particle, the size of the resource particle.

```{r, eval=FALSE}
resIDs = [4,  6, 17, 1]
size =  [100, 87, 156.3, 0.001]
Xpos =   [56, 34, 567, 2]
Ypos =   [876, 98, 32, 45]
```

Above, the particle with ID of "4" has a size of 100 and is located at position x=56, y=876.

*Inert tracers particles*--These are objects that move or flow into and through the environment. Inert tracers only interact with physical barriers. The use of these tracers particles allows for physically dynamic attributes of a system to be quantified. Examples are ecosystem residence time and the ease of passive movement or inertia through the system. Like other individual-level objects in Simplex, tracers are distinguished by corresponding elements of several lists. Each specific position in the list corresponds to the same tracer particle. Information stored about each tracer particle includes time spent in the system, 2D spatial location and particle ID.

@@

**System level state variables**  
Each run of Simplex begins with random choices for the specific structure of a log-series metacommunity and values of width (5 to 100), height (5 to 100), flow rate (0.0 to 1.0 units along the x and y axis per time step), the number, size, and location of physical barriers, the number and direction of environmental gradients, the rate and severity of stochastic disturbance, the rate of fluctuation in basal flow rate, and degrees of fluctuation and syncronization for the inflow of individual organisms and resources, which range from completely in-sync to completely out of sync.

**Spatial and temporal scales**  
\setlength\parindent{36pt}
The two general aspects of scale are grain (i.e., resolution) and extent (e.g., total area). The environment of Simplex models is two dimensional and can vary along each axis from 5 to 100 discrete units. This makes for a potential total extent of 25 to 10,000 discrete patches, each with a grain of 1 square unit. However, all particles move in decimal units, meaning that individual particles can occupy practically infinite locations within patches. Extent of time in Simplex models refers to residence time, i.e., the average amount of time that individual particles spend in the system. As shown by test simulations, residence time for individual particles (organisms, resources, inert tracers) can vary across five orders of magnitude.

Grain is the smallest unit over which change can happen. For example, the original ODD documentation states “One time step represents one year and simulations were run for 100 years. One grid cell represents 1 ha and the model landscape comprised 1,000 x 1,000 ha”. In Simplex, spatial grain is effectively determined by slowest rate at which individuals can traverse the landscape, which can vary from 0.00001% to 10% of the x or y axis in one time step.

@@

*Core simulation process*. Simplex begin simulating immediately after assembling a model from random combinations of state-variables and processes. Simplex models advance turnover of the environmental matrix according to the rate of flow, which can be as low as 0.0 (i.e., an effectively static environment) or as high as 1.0 (i.e., where inert particles will flow 1.0 units of distance unless hindered by barriers). After each iteration of flow, every individual is given the chance to consume, grow, reproduce, starve, die, and to disperse towards environmental optima. However, each of these processes is probabilistic.

*Simulated life history*.\setlength\parindent{36pt} Simplex models simulate growth, reproduction, and death via weighted random sampling. This simulates the partly probabilistic and partly deterministic nature of environmental filtering and individual-level interactions. For example, individuals with a high cell quota have a low probability of dying and a high probability of reproducing.

Inflow/Entrance. Resources and individuals enter from any point in the environment. Species identities of inflowing propagules are chosen at random from a log-series distribution, which often approximates the distribution of abundance among species in ecological communities (Hubbell 2001). Along with a species ID and species-specific maximum rates of resource uptake, active dispersal, resource efficiencies, cell maintenance, and environmental optima, each propagule was given a unique ID and a multi-resource cell quota that represent the state of internal resources. The average of these cell quotas determines the probability of reproduction.

Dispersal. Individuals are allowed to actively move along environmental gradients and even against the direction of environmenta flow and towards their environmental optima. A better match to the environmental optima increases the chance of reproduction and an individual's ability to consume, grow, and reproduce.

Consumption & growth. Sampled individuals consume resources according to their specific maximum rates of uptake and grow according to specific resource efficiencies and maintenance costs. However, can consumption can only happen in Simplex if individuals physically encounter resource particles. Uptake increases the individual cell quotas and decreases the size of the resource particle. Individual cell quotas are then decreased according to a specific physiological maintenance cost.

Reproduction. Reproduction in Simplex is currently limited to clonal reproduction with the possibility of mutation. Individuals reproduce with a probability determined by the combination of mean cell quota and the proportional match to the environmental optima. The cell quota of each resource is evenly divided between the individual and its daughter. The daughter is then given a unique individual ID and the species ID of its mother, i.e., unless in the case of a random speciation event.

@@

**Design concepts**  
*Ecological complexity*. Simplex assembles models from random combinations of constraints (state-variables) and processes to generate output data that allow the user to test the general influence of state-variables and processes.

*Process-based stochasticity*. Simplex was designed to incorporate random sampling and probability-based changes in as many ways as possible. This does not imply neutral dynamics or completely random dynamics, but instead makes all processes and dynamics (even strong environmental filtering) subject to a degree of uncertainty.

*Nutrient limited growth*. All models assembled by Simplex employ the universal concept that individual growth, reproduction, and activity is limited by resources within individuals (endogenous) rather than resouces simply within the environment (exogenous).

@@

These data are stored in .csv files. The file *SimData.csv* is stored an R-formatted data.frames. These files can be directly imported into an R or Python environment.

@@

Droop (1968, 1983) gave a relationship between specific growth rate ($\mu$) and cell quota ($Q$): $$\mu = \mu'_{m} (1 - k_{q}/Q)$$
where $k_{q}$ is the minimum cell quota needed for life, also referred to as the subsistence quota. $\mu'_{m}$ is the theoretical maximum growth rate.

*Maintenance cost of Pirt*. Pirt (1965) states "The variation, with growth rate, of the yield of organism from the substrate used as energy source is attributed to consumption of energy at a constant rate for cell maintenance." He derived a relationship between the growth yield (biomass), the growth rate, and metabolic maintenance. Simplex models use Pirt's concept of a constant maintenance requirement. Simplex also draws from Pirt's simple relation for substrate use: $use(total) = use(maintenance) + use(growth)$

@@

The Simplex platform assembles individual-based models (IBMs) from random combinations of state-variables and processes, to simulate high degrees of variation and complexity. In this way, Simplex allows ecologists to examine whether the influence of particular variables should be robust to a wide range of conditions. For example, if a user wants to examine the influence of area on commonness and rarity, or the potential relationship between growth rate and physiological maintenance, then the user only needs to run Simplex, allow it to assemble and simulate a large number of models, and then analyze the output files. This is unlike many IBMs that are often built for a specific system or for individuals with highly specified behaviors and abilities (e.g. learning).

In addition to its unique modeling aspects, Simplex includes analysis code that shows users how to import and analyze simulated datasets, including simple univariate relationships, descriptive statistics, variance partitioning, and the fitting of species abundance models. By taking advantage of modern computational tools, version control, social coding, Simplex is well-positioned to continue development as both a research and education/instructional tool for ecology.

Future developments and released versions of Simplex will have trophic dynamics including predator-prey, symbionts and mutualists, and producers and consumers. Individuals will also be attributed with additional behaviors (e.g., search patterns), sensing (e.g., of resources), and learning (e.g., predator avoidance). Additionally, individuals will also be attributed with genomes. Initially, these genomes will be represented by sets of sequences of A's, C's, T's, and G's representing genes that can be mutated as they're passed on through clonal reproduction, or mutated and recombined in the case of sexual haploid or diploid reproduction. A precursor to this has already been developed in Locey (2013), the code for which is available here: https://github.com/klocey/SYMENT.

Future developments will also include explicit stoichiometry, where individual resource particles have ratios of Carbon, Nitrogen, and Phosphorus, as well as degrees of biocomplexity. For example, given the chemical structure of phosphate and the size of a resource particle, the particle can be assigned *x* units of phosphorus as well as the biocomplexity value of phosphate, which can be estimated as a form of entropy (Bertz 1983). Likewise, improvements and future versions of Simplex will provide increasing numbers of files for statistical analysis of Simplex output. These files will, in a sense, be able to be used as a complete analysis of some particular aspect of Simplex output (e.g., demographic, resource-related, spatial, diversity-related, multivariate, ecological model testing). These and all Simplex files are freely available and open source, allowing users to modify Simplex source code to suit their particular needs.

@@

Black AJ, McKane AJ (2012) Stochastic formulation of ecological models and their applications. Trends in Ecology and Evolution. 27:337-345.

Bertz, S (1983) On the complexity of graphs and molecules. Bulletin of Mathematical Biology 45:849–855.

DeAngelis DL, Gross LJ (1992) Individual-based Models and Approaches in Ecology: Populations, Communities, and Ecosystems. Chapman & Hall. New York.

Droop MR (1983) Twenty-five years of algal growth kinetics, a personal view. Botanica Marina. 26: 99-112.

Gotelli NJ, Entsminger GL (2001) Swap and fill algorithms in null model analysis: rethinking the knight’s tour. Oecologia. 129:281–291

Grimm V, Railsback SF (2005) Individual-based Modeling and Ecology. Princeton University Press, Princeton.

Grimm V, Berger U, Bastiansen F, Eliassen S, Ginot V, Giske J, Goss-Custard J, Grand T, Heinz SK, Huse G, Huth A, Jepsen JU, Jørgensen C, Mooij WM, Mu¨ller B, Pe’er G, Pioub C, Railsback SF, Robbins AM, Robbins MM, Rossmanith E, Ru¨ger N, Strand E, Souissim S, Stillmane RA, Vabøg R, Visser U, DeAngelis DL (2006) A standard protocol for describing individual-based and agent-based models. Ecological Modeling. 198: 115-126.

Hartl DL, Clark AG (1997) Principle of Population Genetics. Sinauer Associates, Inc. Publishers, Sunderland, MA.

Hubbell SP (2001) The unified neutral theory of biodiversity and biogeography. Princeton University Press, Princeton, NJ.

Leadbeater BSC (2006) The 'Droop equation' - Michael Droop and the legacy of the 'Cell-quota model' of phytoplankton growth. Protist. 157: 345-358.

Locey K (2013) Synthesized ecological molecular neutral theory. figshare. http://dx.doi.org/10.6084/m9.figshare.640105.

Locey KJ, Lennon JT. (2015) Scaling laws predict global microbial diversity. PeerJ PrePrints 3:e1808 https://dx.doi.org/10.7287/peerj.preprints.1451v1

Matthews RB, Gilbert NG, Roach A, Polhill JG, Gotts NM (2007) Agent-based land-use models: a review of applications. Landscape Ecology. 22: 1447-1459.

Pirt SJ (1965) The Maintenance Energy of Bacteria in Growing Cultures. Proceedings of the Royal Society of London. Series B, Biological Sciences, 163:224-231.

Rosindell J, Harmon LJ, Etienne RS (2015) Unifying ecology and macroevolution with individual-based theory. Ecology Letters. 18: 472-482.

Succi S (2001) The Lattice Boltzmann Equation: For Fluid Dynamics and Beyond. Oxford University Press. New York.
